---
title: "Getting started with buckethost"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with buckethost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`buckethost` turns any S3-compatible bucket into a browsable, remotely
readable data repository. This vignette covers day-to-day use; for the full
"how do I get data into a bucket in the first place" recipe see
`vignette("migration-workflow")`.

## Configure once

Functions resolve connection details from options (or `BUCKETHOST_*`
environment variables). Set them once — ideally in your `.Rprofile`:

```{r}
library(buckethost)

options(
  buckethost.endpoint  = "https://object-arbutus.cloud.computecanada.ca",
  buckethost.container = "predictiveecology",
  buckethost.remote    = "arbutus"   # name of the rclone remote
)
```

Three identifiers, easily confused:

- **endpoint** — the storage host, no bucket name.
- **container** — the bucket name.
- **remote** — the *local* name of your `rclone` remote (from `rclone config`).
  Only the mutating functions (`bucketUpload()`, `bucketDelete()`,
  `bucketVerify()`, and uploading indexes) use it.

`bucketBaseUrl()` and `bucketRcloneRemote()` show how these combine:

```{r}
bucketBaseUrl()        #> "https://object-arbutus.cloud.computecanada.ca/predictiveecology"
bucketRcloneRemote()   #> "arbutus:predictiveecology"
```

## Discover what's there

`bucketLs()` reads the bucket's public listing API (no credentials needed for
a public-read bucket) and returns one row per object:

```{r}
objs <- bucketLs(prefix = "SCANFI_v2/1985")
head(objs)
#>                                          key      size                  modified
#> 1 SCANFI_v2/1985/SCANFI_age_median_1985_v2.tif 154238911 2026-01-19T...
```

Build URLs from keys:

```{r}
bucketUrl("SCANFI_v2/1985/SCANFI_age_median_1985_v2.tif")
#> "https://object-arbutus.cloud.computecanada.ca/predictiveecology/SCANFI_v2/1985/..."
```

## Read rasters remotely

`bucketRaster()` wraps `terra::rast("/vsicurl/<url>")`. For a Cloud-Optimized
GeoTIFF, GDAL fetches only the byte ranges it needs:

```{r}
r <- bucketRaster("SCANFI_v2/1985/SCANFI_age_median_1985_v2.tif")
r                       # metadata only; nothing downloaded yet

aoi <- terra::vect(my_study_area)
sub <- terra::crop(r, terra::project(aoi, terra::crs(r)))   # a few hundred KB
terra::plot(sub)
```

## Mutate the bucket

Uploads and deletes wrap `rclone` and fail loudly with rclone's own error
text. rclone only transfers changes, so re-running is cheap.

```{r}
bucketUpload("~/local/SCANFI/1985", "SCANFI_v2/1985",
              filterFile = "~/scanfi.filter",
              extra = c("--retries", "10", "--stats", "60s"))

res <- bucketVerify("~/local/SCANFI/1985", "SCANFI_v2/1985",
                     filterFile = "~/scanfi.filter")
res$ok

bucketDelete("SCANFI_v2/1985/obsolete.tif")          # prompts to confirm
bucketDelete("SCANFI_v2/old", recursive = TRUE,
              confirm = FALSE)                         # whole prefix, no prompt
```

## Generate the browsable catalogue

`generateIndexes()` writes one `index.html` per folder level (sortable
columns, breadcrumbs, a per-page disclaimer block) and uploads them:

```{r}
generateIndexes(
  heading = "PredictiveEcology Temporary Data Repository",
  disclaimerHtml = paste0(
    "<div class='hero'><p><strong>These data are not produced by the ",
    "PredictiveEcology group</strong> and are only hosted here to ease ",
    "open data access.</p></div>"
  ),
  hostNote = "Hosted on the Digital Research Alliance of Canada's Arbutus object storage."
)
```

Browse the result at `<endpoint>/<container>/index.html`. Inspect the pages
without uploading by passing `dryRun = TRUE`, which returns the HTML:

```{r}
pages <- generateIndexes(dryRun = TRUE)
cat(pages[[""]])        # the root index.html
```

To customise the look, copy the shipped template and edit it:

```{r}
file.copy(
  system.file("templates", "index.html", package = "buckethost"),
  "my-template.html"
)
generateIndexes(template = "my-template.html")
```
