---
title: "Migrating data to an object store (the Arbutus recipe)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating data to an object store (the Arbutus recipe)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This is the end-to-end recipe used to move large GeoTIFF datasets from Google
Drive to the Digital Research Alliance of Canada's **Arbutus** object storage,
expose them via `/vsicurl/` for remote-readable access in R (`terra`), and
provide a browsable HTML catalogue with `buckethost`.

It is deliberately specific — future-you debugging a six-month-later problem
will want the exact flags and gotchas, not a sanitised summary. Most of it is
dashboard / shell work done once; the R parts are at the end.

## Why object storage over Google Drive?

- Google Drive doesn't support HTTP range requests reliably, so partial reads
  with `terra::rast("/vsicurl/...")` don't work — every read pulls the whole
  file.
- Arbutus is S3/Swift-compatible, Canadian-hosted, and free for academic PIs
  (Rapid Access Service allocation).
- Once files are Cloud-Optimized GeoTIFFs (COGs) on Arbutus,
  `terra::rast()` + `crop()` transfers only the bytes needed for the area of
  interest — typically a few hundred KB per query.

## Why a fast transfer node, not your usual login host?

A government-network host gave ~1 MB/s sustained during testing; a 321 GB
migration would have taken days. The Alliance's **Fir** cluster is on the
CANARIE research network with ~600+ MB/s to both Drive and Arbutus — the same
321 GB completes in minutes. Run the transfer there.

## Phase 0 — accounts and credentials (one-time, dashboard)

**A. CCDB account active.** At <https://ccdb.alliancecan.ca>: confirm PI status
and active allocations; under *My Account → Access Systems* enable the cluster
you'll transfer from (e.g. Fir); under *Multifactor Authentication Management*
enroll a Duo device and save 10 backup codes.

**B. Arbutus cloud project + storage.** Request via the **RAS** (Rapid Access
Service) form on CCDB. The form requires non-zero VM resources even for
object-storage-only use, so choose the minimum (1 vCPU, 1 instance, 1 volume,
1.5 GB RAM, 1 floating IP, 20 GB persistent). The number that matters is
**Object Storage** — request what you need. Justify: hosting COG outputs for
collaborator access via `/vsicurl/`; HPC clusters don't fit (they're for
compute, not public HTTPS serving); include a maintenance plan. Turnaround is
typically a week or two.

**C. EC2 (S3-compatible) credentials.** The OpenStack *Application
Credentials* in the dashboard are **not** what you want — those use
Swift-native auth, not S3. The Horizon dashboard removed the direct "EC2
Credentials" button, so use the CLI:

```bash
pipx install python-openstackclient        # or pip install --user on Windows
source ~/Downloads/<project>-openrc.sh
openstack ec2 credentials create
```

This needs port 5000 (Keystone), which is **blocked on many government
networks**. Workaround: run it once from a home network or cellular hotspot.
The resulting EC2 access key + secret are portable — once created they work
from anywhere over port 443. Save them to a password manager.

**D. Public container.** In the Arbutus dashboard: *Project → Object Store →
Containers → Create container* named e.g. `predictiveecology`. Set its access
to **Public** so `/vsicurl/` reads work without authentication.

## Phase 1 — configure rclone on the transfer node

Fir ships `rclone` (verify with `rclone version`). You need two remotes:
`arbutus` (S3) and `gdrive` (Google Drive source).

### 1a. The `arbutus` S3 remote

`rclone config` → `n` (new) → name `arbutus` → storage `s3` → provider
`Other` → `env_auth` false → paste your EC2 access key / secret → region blank
→ endpoint `https://object-arbutus.cloud.computecanada.ca` →
location_constraint blank → acl `private` → **Edit advanced config: yes**, and
set the one critical option:

```
force_path_style = true
```

Arbutus's Swift S3 gateway doesn't support virtual-hosted-style URLs
(`bucket.endpoint.com`). Without `force_path_style = true`, every request
fails with signature or DNS errors. Accept defaults for everything else. The
resulting `~/.config/rclone/rclone.conf` should contain:

```ini
[arbutus]
type = s3
provider = Other
access_key_id = <key>
secret_access_key = <secret>
endpoint = https://object-arbutus.cloud.computecanada.ca
acl = private
force_path_style = true
```

Test: `rclone lsd arbutus:` should list your container(s) with no error.

### 1b. The `gdrive` remote

Google Drive OAuth needs a browser, but the node is headless. Bridge the
callback with SSH port forwarding from your laptop:

```bash
ssh -L 53682:localhost:53682 <user>@fir.alliancecan.ca
```

In *that same session*: `rclone config` → `n` → name `gdrive` → storage
`drive` → client_id/secret blank → scope `drive.readonly` (safer for a
one-way migration) → service_account_file blank → advanced config `n` → "Use
web browser to automatically authenticate?" `y`. rclone prints a
`http://127.0.0.1:53682/auth?state=...` URL — paste it into your laptop
browser; the SSH tunnel forwards it to rclone waiting on Fir. Authorize, then
"Configure as Shared Drive?" `n` (unless it's a Team Drive) → confirm.

Test: `rclone lsd gdrive:` lists your Drive folders. If your data is under
"Shared with me", either add a My Drive shortcut to it, or add
`--drive-shared-with-me` to commands targeting that path.

## Phase 2 — migrate

Work inside `tmux` so the transfer survives disconnects: `tmux new -s migrate`
(detach with `Ctrl-b d`, reattach with `tmux attach -t migrate`).

**2a. Filter file** (explicit top-to-bottom ordering, first match wins):

```bash
cat > ~/scanfi.filter << 'EOF'
- *.tif.tiff
+ *.tif
+ *.tif.ovr
+ *.tif.aux.xml
- *
EOF
```

**2b. Run the copy**, looping over subfolders. Note the destination uses
underscores (`SCANFI_v2`) instead of the source's space for cleaner URLs:

```bash
for year in 1985 1990 1995 2000 2005 2010 2015 2020 2025; do
  echo "=== $(date) Starting $year ==="
  rclone copy \
    "gdrive:Shared/Hosted/SCANFI/SCANFI v2/$year" \
    "arbutus:predictiveecology/SCANFI_v2/$year" \
    --filter-from ~/scanfi.filter \
    --transfers 16 --timeout 5m --contimeout 60s \
    --retries 10 --low-level-retries 20 \
    --stats 60s --log-file ~/rclone_scanfi_$year.log --log-level INFO
  echo "=== $(date) Completed $year ==="
done
```

**2c. Verify** counts/sizes, then a size-only check against the source:

```bash
for year in 1985 1990 1995 2000 2005 2010 2015 2020 2025; do
  rclone size arbutus:predictiveecology/SCANFI_v2/$year
done

rclone check \
  "gdrive:Shared/Hosted/SCANFI/SCANFI v2/1985" \
  "arbutus:predictiveecology/SCANFI_v2/1985" \
  --filter-from ~/scanfi.filter --size-only
```

The equivalent from R, once `buckethost` is configured:

```{r}
bucketVerify("gdrive-or-local/1985", "SCANFI_v2/1985",
              filterFile = "~/scanfi.filter")$ok
```

## Phase 3 — generate the catalogue

From any R session with the `arbutus` rclone remote configured:

```{r}
library(buckethost)
options(
  buckethost.endpoint  = "https://object-arbutus.cloud.computecanada.ca",
  buckethost.container = "predictiveecology",
  buckethost.remote    = "arbutus"
)

generateIndexes(
  heading = "PredictiveEcology Temporary Data Repository",
  disclaimerHtml = paste0(
    "<div class='hero'>",
    "<p><strong>These data are not produced by the PredictiveEcology group",
    "</strong> and are only hosted here to ease open data access.</p>",
    "<p>For SCANFI v2 data, see the authoritative source: ",
    "<a href='https://open.canada.ca/data/en/dataset/",
    "07653869-f303-46c2-a04e-9ab479b73cbf' target='_blank' rel='noopener'>",
    "open.canada.ca</a></p></div>"
  ),
  hostNote = "Hosted on the Digital Research Alliance of Canada's Arbutus object storage."
)
```

Browse it at
`https://object-arbutus.cloud.computecanada.ca/predictiveecology/index.html`.
Re-run `generateIndexes()` any time data is added or changed.

## Phase 4 — the actual payoff: read it from anywhere

```{r}
library(terra)
r <- bucketRaster("SCANFI_v2/1985/SCANFI_age_median_1985_v2.tif")
r                                            # metadata only

b   <- SpaDES.tools::randomStudyArea(size = 1e9, seed = 123)
sub <- crop(r, project(vect(b), crs(r)))     # only the AOI's bytes transfer
plot(sub)
```

## Troubleshooting FAQ

- **Every S3 request fails with signature/DNS errors** → you forgot
  `force_path_style = true` on the rclone S3 remote.
- **`openstack ec2 credentials create` hangs/refuses** → port 5000 is blocked
  on your network. Run it once from home or a hotspot; the credentials are
  portable afterwards.
- **OAuth callback never completes on a headless node** → the SSH
  `-L 53682:localhost:53682` tunnel isn't up, or you ran `rclone config` in a
  different session than the one with the tunnel.
- **Application Credentials don't work with rclone** → those are Swift-native;
  you need **EC2** credentials for the S3 protocol.
- **Data is in "Shared with me" and rclone can't see it** → add a My Drive
  shortcut, or use `--drive-shared-with-me`.
